\chapter{Writing Systems} \label{writing_systems}

% \section{Introduction} % \label{introduction}

Writing systems arise and develop in a complex mixture of cultural,
technological and practical pressures. They tend to be highly conservative, in
that people who have learned to read and write in a specific way (however
impractical or tedious) are mostly unwilling to change their habits, e.g.~they
tend to resist spelling reforms. In all literate societies there exists a strong
socio-political mainstream that tries to force unification of writing (for
example by strongly enforcing ``right'' from ``wrong'' writing in schools).
However, there is also a large community of users who take as many liberties in
their writing as they can get away with.

For example, the writing of tone diacritics in Yoruba is often proclaimed to be
the right way to write, although many users of Yoruba writing seem to be
perfectly fine with leaving them out. As pointed out by the proponents of the
official rules, there are some homographs when leaving out the tone diacritics
\citet[44]{Olumuyiw2013}. However, writing systems (and the languages they
represent) are normally full of homophones, which is normally not a problem at
all for speakers of the language. More importantly, writing is not just a purely
functional tool, but just as importantly it is a mechanism to signal social
affiliation. By showing that you \textit{know the rules} of expressing yourself
in writing, others will more easily accept you as a worthy participant in their
group. And that just as well holds for obeying to the official rules when
writing a job application, as for obeying to the informal rules when writing an
SMS to classmates in school. The case of Yoruba writing is an exemplary case, as
even after more than a century of efforts to standardize the writing systems,
there is still a wide range of variation in daily use \citet{Olumuyiw2013}.

The sometimes cumbersome and sometimes illogical structure, and the enormous
variability of existing writing systems is a fact of life scholars have to
accept and should try to adapt to as good as possible. Our plea here is a
proposal for a formalization to do exactly that.

When considering the worldwide linguistic diversity, including all
lesser-studied and endangered languages, there exist numerous different
orthographies using symbols from the same scripts. For example, there are
hundreds of orthographies using Latin-based alphabetic scripts. All of these
orthographies use the same symbols, but these symbols differ in meaning and
usage throughout the various orthographies. To be able to computationally use
and compare different orthographies, we need a way to specify all orthographic
idiosyncrasies in a computer-readable format (a process called
\textsc{tailoring} in Unicode parlance). We call such specifications
\textsc{orthography profiles}. Ideally, these specifications have to be
integrated into so-called Unicode locale descriptions, though we will argue that
in practice this is often not the most useful solution for the kind of problems
arising in the daily practice of linguistics. Consequently, a central goal of
this paper is to flesh out the linguistic challenges for locale descriptions,
and work out suggestions to improve their structure for usage in a linguistic
context. Conversely, we also aim to improve linguists' understanding and
appreciation for the accomplishments of the Unicode Consortium in the
development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, in our
current research our goal is to develop quantitative methods for language
comparison and historical analysis in order to investigate worldwide linguistic
variation and to model the historical and areal processes that underlie
linguistic diversity,
cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the \textit{same} language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme
separation of strings of symbols, a process we call \textsc{grapheme
tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words,
sentences) in a specific source, our first processing step is to specify how
these strings have to be separated into graphemes, considering the specific
orthographic conventions used in a particular source document. Our experience is
that such a graphemic tokenization can be performed without extensive in-depth
knowledge about the phonetic and phonological details of the language in
question. For example, the specification that $<$ou$>$ is a grapheme of English
is a much easier task than to specify what exactly the phonetic values of this
grapheme are in any specific occurrence in English words. Grapheme separation is
a task that can be performed relatively reliably and with limited availability
of time and resources (compare, for example, the task of creating a complete
phonetic or phonological normalization).

Although grapheme tokenization is only one part of the solution, it is an
important and highly fruitful processing step. Given a grapheme tokenization,
various subsequent tasks become easier, like (a) temporarily reducing the
orthography in a processing pipeline, e.g.~only distinguishing high versus low
vowels; (b) normalizing orthographies across sources (often including temporary
reduction of oppositions), e.g.~specifying an (approximate) mapping to the
International Phonetic Alphabet; (c) using co-occurrence statistics across
different languages (or different sources in the same language) to estimate the
probability of grapheme matches, e.g.~with the goal to find regular sound
changes between related languages or transliterations between different sources
in the same language.

Before we deal with these proposals, in the first part of this paper (Sections
\ref{encoding} through \ref{ipa-meets-unicode}) we give an extended introduction
to the notion of encoding (Section \ref{encoding}) and writing systems, both
from a linguistic perspective and from the perspective of the Unicode Consortium
(Section \ref{terminology}). We consider the Unicode Standard to be a
breakthrough (and ongoing) development that fundamentally changed the way we
look at writing systems, and we aim to provide here a slightly more in-depth
survey of the many techniques that are available in the standard. A good
appreciation for the solutions that the Unicode Standard also allows for a
thorough understanding of possible pitfalls that one might encounter when using
it (Section \ref{unicode-pitfalls}). As an example of the current
state-of-the-art, we discuss the rather problematic marriage of the
International Phonetic Alphabet (IPA) with the Unicode Standard (Section
\ref{ipa-meets-unicode}).

The second part of the paper (Sections \ref{orthography-profiles} and
\ref{use-cases}) describes our proposals for how to deal with the Unicode
Standard in the daily practice of (comparative) linguists. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Section \ref{orthography-profiles}). Finally, we discuss practical
issues with actual examples (Section \ref{use-cases}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries.

The following conventions are followed in this paper. All phonemic and phonetic
representations are given in the International Phonetic Alphabet (IPA), unless
noted otherwise \citep{IPA2005}. Standard conventions are used for
distinguishing between graphemic < >, phonemic / / and phonetic [ ]
representations. For character descriptions, we follow the notational
conventions of the Unicode Standard \citep{Unicode2014}. Character names are
represented in small capital letters (e.g.~\textsc{latin small letter schwa})
and code points are expressed as U\emph{+n} where \emph{n} is a four to six
digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
