\chapter{Testing code inclusion}


Trying to include code through knitr. 
Different languages can be easily 
included!

There is a bug for non-R engines: the fontsize settings and line spaceing are not correct. I have 
filed an issue, hopefully this will be fixed soon. 

For syntax colouring, install the package `highlight' and for nice tables 
install the package `xtable'.

<<bash-test, engine='bash'>>=
echo 'this is a simple bash example'
ls | wc
@

<<engine='bash', echo = FALSE>>=
echo -e 'note that bash chunks are executed as one, with all output at the end.
You will have to make different chunks to get output separated.
And add manual linebreaks!'
@

Spacing of lines is fine. I think without background looks better

<<>>=
# adding comments inside code? Probably not a good idea
(example <- 'this is a simple R example')

test = 3+4
test
@

<<>>=
library(qlcData) 
profile = write.profile('AΑА')
@

<<Rtest2, tidy = TRUE, results = 'asis', echo = FALSE>>=
print(xtable(profile, caption = 'test', label = 'bla')
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'small'
  , table.placement = 'H'
  )
@



You can refer to variables by using \textbackslash Sexpr{}. 
For example, the length of the example string is \Sexpr{nchar(example)}.
Crossreferencing also works, see Table~\ref{bla}.

<<python-test, engine='python'>>=
print('bla' + 'bla')
print(3+4)
@

You have to add the option ``engine.path='python3'' to get Python 3.

<<python3-test, engine='python', engine.path='python3'>>=
print('python3 needs an extra tag')
uni = 'aɽɮz'
print(uni)
@

Testing: does referencing to variables also work in Python? It does give an error, so no... !